\documentclass{ximera}
\input{../../../preamble.tex}

\author{Zack Reed}
%borrowed from selinger linear algebra
\title{Cross Product}
\begin{document}
\begin{abstract}

\end{abstract}
\maketitle

\section*{Cross Product and its Properties}

 Here, we introduce the cross product, which has many applications in physics and engineering.  It also has important geometric properties.
 
\subsection*{Motivating the Cross Product}

Let's return to the unfinished problem of the previous section:

\begin{example}\name{Normal and standard equations}\label{exa:normal-from-three-points}

    Find normal and standard equations for the plane through the points
    $P = (0,1,3)$, $Q=(2,-1,0)$, and $R=(1,2,2)$.
  \end{example}
  
  \begin{solution}
    The goal is to find the normal vector to the plane so that we can more explicitly define the plane. For any vector $\vec{v}$, there is a nice linear transformation defined only on $\RR^3$, called the cross product and denoted $\vec{v}\times$, which will for any vector $\vec{u}$ give a third vector $\vec{w}$ that is orthogonal to both $\vec{v}$ and $\vec{u}$.

    Much like in $\RR^2$, where if you have a vector $\vec{v}=\begin{bmatrix}
      a\\b
    \end{bmatrix}$, you can immediately calculate a vector orthogonal to it by $\vec{n}=\begin{bmatrix}
      -b\\a
    \end{bmatrix}$, we have a similar formulation in $\RR^3$.

    In $\RR^3$, if you have two vectors $\vec{u}=\begin{bmatrix}
      a_1\\a_2\\a_3
    \end{bmatrix}$ and $\vec{v}=\begin{bmatrix}
      b_1\\b_2\\b_3
    \end{bmatrix}$, you can explicitly find a vector orthogonal to both $\vec{u}$ and $\vec{v}$ (so long as the two are not on the same line) by the vector

    $$\vec{w}=\begin{bmatrix}
      a_2b_3-a_3b_2\\a_3b_1-a_1b_3\\a_1b_2-b_1a_2
    \end{bmatrix}$$

    This vector $\vec{w}=\vec{v}\times\vec{u}$ is exactly what we need to describe any plane from two vectors, so let's test it out on our example plane and determine its normal equation!

    First, let's find the two direction vectors and point that define the plane. If we center at $P$, we can use  $\longvect{PQ}$ and $\longvect{PR}$ as the planar direction vectors. 

    From the last section, we determined that the direction vectors are $\longvect{PQ}=\startmat{r} 2 \\ -2 \\ -3 \stopmat$ and $\longvect{PR}=\startmat{r} 1 \\ 1 \\ -1 \stopmat$.
    
    Using the general formula above, we can hopefully find a normal vector for the
    plane:
    \begin{equation*}
      \vect{n}
      ~=~
      \longvect{PQ} \times \longvect{PR}
      ~=~
      \startmat{r} 2 \\ -2 \\ -3 \stopmat
      \times
      \startmat{r} 1 \\ 1 \\ -1 \stopmat
      ~=~
      \startmat{r} \answer{5} \\ \answer{-1} \\ \answer{4} \stopmat.
    \end{equation*}
    \begin{center}
      \begin{tikzpicture}[rotate=-10]
        \filldraw[draw=red!80,fill=red!10](-3,0,-3.5) -- (3,0,-3.5) -- (3,0,3.5) -- (-3,0,3.5) -- cycle;
        \draw[->,thick,blue!80!black](0,0,0) -- node[left, pos=0.4] {$\vect{n}$} (100:2);
        \draw[->,thick,blue!80!black](0,0,0) -- (2,0,1);
        \draw[->,thick,blue!80!black](0,0,0) -- (1,0,-2);
        \fill (0,0,0) circle [radius=2.2pt] node [left=3pt] {$P$};
        \fill (2,0,1) circle [radius=2.2pt] node [right=3pt] {$Q$};
        \fill (1,0,-2) circle [radius=2.2pt] node [right=3pt] {$R$};
      \end{tikzpicture}
    \end{center}

    Let's verify that $\vec{n}$ is actually a normal vector for the plane.

    Using the dot product, we get

    $$\startmat{r} 2 \\ -2 \\ -3 \stopmat\cdot \startmat{r} \answer{5} \\ \answer{-1} \\ \answer{4} \stopmat=\answer{10}+\answer{2}+\answer{-12}=0,$$

    and $$\startmat{r} 1 \\ 1 \\ -1 \stopmat\cdot \startmat{r} \answer{5} \\ \answer{-1} \\ \answer{4} \stopmat=\answer{5}+\answer{-1}+\answer{-4}=0,$$

    so indeed, $\vec{n}$ is normal to the plane!

 

    Now we can easily obtain the normal equation from any point on the
    plane (say $P$) and the normal vector we just calculated:
    \begin{equation*}
      \startmat{r}\answer{5} \\ \answer{-1} \\ \answer{4} \stopmat
      \cdot
      \startmat{r} x \\ y-1 \\ z-3 \stopmat
      =
      0
    \end{equation*}




    We get the standard equation by computing the dot products on the
    left- and right-hand sides:
    \begin{equation*}
      \answer{5}x +\answer{-1} y + \answer{4}z = 11.
    \end{equation*}

  \end{solution}

  More than just in this one example, the calculation 

  
  $$\vec{w}=\begin{bmatrix}
      a_2b_3-a_3b_2\\a_3b_1-a_1b_3\\a_1b_2-b_1a_2
    \end{bmatrix}$$

  will yield the vector in $\RR^3$ orthogonal to any pair of vectors in $\RR^3$ (that do not lie on the same line) $\vec{u}=\begin{bmatrix}
      a_1\\a_2\\a_3
    \end{bmatrix}$ and $\vec{v}=\begin{bmatrix}
      b_1\\b_2\\b_3
    \end{bmatrix}$.

   

  We will primarily use MATLAB to calculate this for us, but it is also beneficial to have on hand a few ways to calculate this orhtogonal vector, as in the following definition:
 
\begin{definition}\label{def:crossproduct} Let $\vec{u=\begin{bmatrix}u_1\\u_2\\u_3\end{bmatrix}}$ and $\vec{v}=\begin{bmatrix}v_1\\v_2\\v_3\end{bmatrix}$ be vectors in $\RR^3$.  The \dfn{cross product} of $\vec{u}$ and $\vec{v}$, denoted by $\vec{u}\times\vec{v}$, is given by
\begin{align*}
\vec{u}\times\vec{v}&=(u_2v_3-u_3v_2)\vec{i}-(u_1v_3-u_3v_1)\vec{j}+(u_1v_2-u_2v_1)\vec{k} \\
&=(u_2v_3-u_3v_2)\begin{bmatrix}1\\0\\0\end{bmatrix}-(u_1v_3-u_3v_1)\begin{bmatrix}0\\1\\0\end{bmatrix}+(u_1v_2-u_2v_1)\begin{bmatrix}0\\0\\1\end{bmatrix}=\begin{bmatrix}u_2v_3-u_3v_2\\-u_1v_3+u_3v_1\\u_1v_2-u_2v_1\end{bmatrix}
\end{align*}

If we fix a vector $\vec{u}$, then we get a linear transformation $\vec{u}\times$ that gives the product with any vector $\vec{v}$ defined by the following matrix

$$\vec{u}\times=\begin{bmatrix}
  0 &-u_3& u_2\\u_3 &0 &-u_1\\-u_2 &u_1 &0
\end{bmatrix}.$$
\end{definition}


 
\begin{example}\label{ex:crossproduct}
Find the cross product of $\vec{u}=\begin{bmatrix}3\\ -10\\ 2\end{bmatrix}$ and $\vec{v}=\begin{bmatrix}-2\\ 4\\ 7\end{bmatrix}$.

\begin{solution}

  (Hint: Use matrix multiplication in MATLAB to speed up the final calculation)

$$\vec{u}\times \vec{v}=\begin{bmatrix}3\\ -10\\ 2\end{bmatrix}\times\vec{v}=\begin{bmatrix}-2\\ 4\\ 7\end{bmatrix}=\begin{bmatrix}\answer{0} & \answer{-2} & \answer{-10} \\
\answer{2} & \answer{0} & \answer{-3} \\
\answer{10} & \answer{3} & \answer{0}
\end{bmatrix}
\begin{bmatrix}
\answer{-2} \\ \answer{4} \\ \answer{7}
\end{bmatrix}
=
\begin{bmatrix}
\answer{-78} \\ \answer{-25} \\ \answer{-8}
\end{bmatrix}$$

\end{solution}
\end{example}

In MATLAB, if you define the vectors $\vec{u}$ and $\vec{v}$, then the command \texttt{cross(u,v)} will calculate the cross product for you.
 
\subsection*{Properties of the Cross Product}
 
% In the next example, we will use vectors $\vec{u}$ and $\vec{v}$ of Example \ref{ex:crossproduct}, and compute $\vec{v}\times\vec{u}$ to illustrate that the cross product is not commutative and to find a relationship between $\vec{u}\times\vec{v}$ and $\vec{v}\times \vec{u}$.  If you have already studied the effect that switching two rows of a matrix has on its determinant, you should be able to guess the outcome of the upcoming computation.
 
\begin{exploration}\label{init:crossproduct2}
What would happen if we took the cross product of the vectors in Example \ref{ex:crossproduct} but reversed the order?
 
Let $\vec{v}=\begin{bmatrix}-2\\ 4\\ 7\end{bmatrix}$ and $\vec{u}=\begin{bmatrix}3\\ -10\\ 2\end{bmatrix}$.

Using MATLAB, we find 

$$
\vec{v}\times \vec{u}
=\begin{bmatrix}\answer{78}\\ \answer{25}\\ \answer{8}\end{bmatrix}=\answer{-1}(\vec{u}\times\vec{v})$$
 
This computation suggests that switching the order of the vectors changes the sign of the result.
\end{exploration}
 
\begin{theorem}\label{th:corssuvnegcrossvu}
Let $\vec{u}$ and $\vec{v}$ be vectors in $\RR^3$, then
\begin{enumerate}
\item $$\vec{u}\times\vec{v}=-(\vec{v}\times\vec{u})$$
\item\label{item:scalarassocofcrossprod} 
$$(k\vec{u})\times \vec{v}=\vec{u}\times (k\vec{v})=k(\vec{u}\times \vec{v})$$
\item\label{item:distofrossprod} 
$$(\vec{u}+\vec{v})\times \vec{w}=\vec{u}\times \vec{w}+\vec{v}\times \vec{w}$$
$$\vec{u}\times (\vec{v}+\vec{w})=\vec{u}\times \vec{v}+\vec{u}\times \vec{w}$$
\end{enumerate}
\end{theorem}
 
The cross product has several important geometric properties. The following problems give us a glimpse of these properties.
 
\begin{exploration}\label{init:ijkcrossproducts}
Compute the following products:
$$\vec{i}\times\vec{j}=\begin{bmatrix}\answer{0}\\\answer{0}\\\answer{1}\end{bmatrix},\quad\vec{j}\times\vec{k}=\begin{bmatrix}\answer{1}\\\answer{0}\\\answer{0}\end{bmatrix}$$
 
\end{exploration}
 
\subsubsection*{Cross Product and the Angle between Vectors}
Recall that the dot product of $\vec{u}$ and $\vec{v}$ is related to the angle $\theta$ between $\vec{u}$ and $\vec{v}$ by the following formula
\begin{align}\label{eq:dotproductformula}\vec{u}\dotp\vec{v}=\norm{\vec{u}}\norm{\vec{v}}\cos\theta\end{align}

We have a very similar property for the cross product.
 
\begin{theorem}\label{th:crossproductsin}
Let $\vec{u}$ and $\vec{v}$ be vectors in $\RR^3$. Let $\theta$ be the angle between $\vec{u}$ and $\vec{v}$ such that $0\leq\theta\leq \pi$. Then
$$\norm{\vec{u}\times\vec{v}}=\norm{\vec{u}}\norm{\vec{v}}\sin\theta$$
\end{theorem}

\subsection*{Vector Products, Areas, Volumes}
 
We conclude by bringing together both the dot and cross products to generate measures of area and volume in $3$-dimensions.

Consider the parallelogram determined by vectors $\vec{u}$ and $\vec{v}$ in $\RR^3$.
 
\begin{center}
\begin{tikzpicture}[scale=1.4]
 
  \filldraw[blue, opacity=0.3](0,0)--(4,2)--(5,4)--(1,2)--cycle;
 
\draw[line width=2pt,red,-stealth](0,0)--(4,2) node[below right]{$\vec{u}$};
  
  \draw[line width=2pt,blue, -stealth](0,0)--(1,2) node[above left]{$\vec{v}$};
\end{tikzpicture}
\end{center}
 
Recall that the area of a parallelogram is given by the product of the length of the base and the height.
As shown in the diagram below, the length of the base is the magnitude of $\vec{u}$. The height, $h$, can be found using trigonometry $$h=\norm{\vec{v}}\sin\theta$$

If we combine this with the base length, $\norm{\vec{u}}$, we exactly get the cross product!

\begin{center}
\begin{tikzpicture}[scale=1.4]
 
  \filldraw[blue, opacity=0.3](0,0)--(4,2)--(5,4)--(1,2)--cycle;
 
\draw[line width=0.5pt, dashed](1,2)--(1.6,0.8);
\node[] at (2, 1.6)   (b) {$h=\norm{\vec{v}}\sin\theta$};
\draw[line width=2pt,red](0,0)--(4,2);
 \node[red] at (2, 0.6)   (b) {$\norm{\vec{u}}$};
 \node[blue] at (0.2, 1.2)   (b) {$\norm{\vec{v}}$};
 \node[] at (0.3, 0.3)   (b) {$\theta$};
  \draw[line width=2pt,blue](0,0)--(1,2);
\end{tikzpicture}
\end{center}
The area of a parallelogram is thus
$$\mbox{Area}=(\text{base})(\text{height})=\norm{\vec{u}}h=\norm{\vec{u}}\norm{\vec{v}}\sin\theta=\norm{\vec{u}\times\vec{v}}$$
We have established the following formula.
 
\begin{formula}\label{form:areaofparallelogram} The area of a parallelogram determined by vectors $\vec{u}$ and $\vec{v}$ in $\RR^3$ is given by
$$\mbox{Area}=\norm{\vec{u}\times\vec{v}}$$
\end{formula}
 
\begin{example}\label{ex:areaOfParFormula}
    Use Formula \ref{form:areaofparallelogram} to find the area of a parallelogram determined by vectors
    $$\vec{u}=\begin{bmatrix}-2\\1\\3\end{bmatrix},\quad \vec{v}=\begin{bmatrix}1\\6\\2\end{bmatrix}$$
\begin{explanation}
    We can start by visualizeing the parallelogram in GeoGebra.  RIGHT-CLICK and DRAG to rotate the image below.  The area of the parallelogram, rounded to two decimal places, is displayed inside the parallelogram.
 
 
\begin{onlineOnly}   
\begin{center}
\geogebra{g7g6kjqm}{600}{400}
\end{center}
\end{onlineOnly}
 
    To find the exact area we compute
    $$\vec{u}\times \vec{v}=\begin{bmatrix}-2\\1\\3\end{bmatrix}\times \begin{bmatrix}1\\6\\2\end{bmatrix}=\begin{bmatrix}\answer{-16}\\\answer{7}\\\answer{-13}\end{bmatrix}$$
 
    $$\mbox{Area}=\norm{\vec{u}\times\vec{v}}=\sqrt{\answer{474}}$$
\end{explanation}
\end{example}
 
Formula \ref{form:areaofparallelogram} can be easily adapted to parallelograms determined by vectors in $\RR^2$, as illustrated by the following example.
 
\begin{example}\label{ex:areaofparallelogram}
Find the area of the parallelogram in the diagram.
\begin{center}
\begin{tikzpicture}[scale=1]
\draw[thin,gray!40] (-3,-1) grid (4,4);
  \draw[<->] (-3,0)--(4,0);
  \draw[<->] (0,-1)--(0,4);
   
  \filldraw[blue, opacity=0.3](0,0)--(-2,2)--(2,4)--(4,2)--cycle;
 
\draw[line width=2pt,red,-stealth](0,0)--(4,2);
 
 
 \draw[line width=2pt,blue,-stealth](0,0)--(-2,2);
  
\end{tikzpicture}
\end{center}
\begin{explanation}
The vectors that determine the parallelogram are
$$\begin{bmatrix}4\\2\end{bmatrix}\quad\text{and}\quad\begin{bmatrix}-2\\2\end{bmatrix}$$
The problem we run into is that these vectors are in $\RR^2$, whereas the cross product is defined only for vectors in $\RR^3$.  We will get around this difficulty by ``padding" our vectors with zeros on the bottom.  In other words, we will consider them as vectors sitting in the $xy$-coordinate plane in $\RR^3$.  This allows us to compute the cross product

$$\begin{bmatrix}4\\2\\0\end{bmatrix}\times\begin{bmatrix}-2\\2\\0\end{bmatrix}=\begin{bmatrix}\answer{0}\\\answer{0}\\\answer{12}\end{bmatrix}$$

The area of the parallelogram is then given by
$$\mbox{Area}=\norm{\begin{bmatrix}\answer{0}\\\answer{0}\\\answer{12}\end{bmatrix}}=\answer{12}$$
\end{explanation}
\end{example}
 
 
\subsection*{The Volume of a Parallelepiped}
 
Our next goal is to find the volume of a three-dimensional figure called a \dfn{parallelepiped}.  A parallelepiped is a six-faced figure whose opposite faces are congruent parallelograms located in parallel planes.  A parallelepiped is a three-dimensional counterpart of a parallelogram, and is determined by three non-coplanar vectors in $\RR^3$.  The figure below shows a parallelepiped determined by three vectors.
 
\begin{center}
\tdplotsetmaincoords{70}{130}
\begin{tikzpicture}
    \draw[->](-2,0,0)--(5,0,0) node[below left]{$y$};
    \draw[->](0,-2,0)--(0,6,0) node[below left]{$z$};
    \draw[->](0,0,-2)--(0,0,5) node[below left]{$x$};
     
    \draw[line width=0.5pt, dashed](3,5,1)--(5,1,-2);
    \draw[line width=0.5pt, dashed](3,5,1)--(-2,4,3);
    \draw[line width=0.5pt, dashed](3,5,1)--(7,9,6);
     
    \filldraw[blue, opacity=0.4] (0,0,0)--(-2, 4, 3)--(2,8,8)--(4,4,5)--cycle;
    \filldraw[blue, opacity=0.5] (2,8,8)--(4,4,5)--(9,5,3)--(7,9,6)--cycle;
    \filldraw[blue, opacity=0.6] (0,0,0)--(4,4,5)--(9,5,3)--(5,1,-2)--cycle;
     
    \draw[->, line width=2pt,blue, -stealth](0,0,0)--(-2,4,3);
    \draw[->, line width=2pt,red, -stealth](0,0,0)--(4,4,5);
    \draw[->, line width=2pt, -stealth](0,0,0)--(5,1,-2);
     
\end{tikzpicture}
\end{center}
 
Consider a parallelepiped determined by vectors $\vec{u}$, $\vec{v}$ and $\vec{w}$, as shown below. 
 
\begin{center}
\tdplotsetmaincoords{70}{130}
\begin{tikzpicture}
    \draw[->, line width=2pt, -stealth,orange,dashed](0,0,0)--(4,0,1)node[above left]{$\vec{v}$};
     
    \draw[line width=0.5pt, dashed](4,0,1)--(5,0,6);
    \draw[line width=0.5pt, dashed](4,0,1)--(9,5,2);
     
    \draw[line width=2pt, gray](1,0,5)--(6,5,6);
    \draw[line width=2pt, gray](10,5,7)--(6,5,6);
    \draw[line width=2pt, gray](10,5,7)--(5,0,6);
    \draw[line width=2pt, gray](1,0,5)--(5,0,6);
    \draw[line width=2pt, gray](6,5,6)--(5,5,1);
    \draw[line width=2pt, gray](9,5,2)--(5,5,1);
    \draw[line width=2pt, gray](10,5,7)--(9,5,2);
     
    \draw[->, line width=2pt,red, -stealth](0,0,0)--(1,0,5)node[below left]{$\vec{u}$};
    \draw[->, line width=2pt, blue, -stealth](0,0,0)--(5,5,1)node[above left]{$\vec{w}$} ;
   
\end{tikzpicture}
\end{center}
 
The volume of a parallelepiped is given by
$$\mbox{Volume}=(\text{area of base})(\text{height})$$
We will consider the parallelogram determined by $\vec{u}$ and $\vec{v}$ to be the base of the parallelepiped.  Thus, the area of the base is given by
$$\mbox{Area of Base}=\norm{\vec{u}\times\vec{v}}$$
\begin{center}
\tdplotsetmaincoords{70}{130}
\begin{tikzpicture}
    \draw[->, line width=2pt, -stealth,orange,dashed](0,0,0)--(4,0,1)node[above left]{$\vec{v}$};
     
    \draw[line width=0.5pt, dashed](4,0,1)--(5,0,6);
    \draw[line width=0.5pt, dashed](4,0,1)--(9,5,2);
     
    \draw[line width=0.5pt, dashed](5,5,1)--(0,5,0);
     
    \draw[line width=2pt, gray](1,0,5)--(6,5,6);
    \draw[line width=2pt, gray](10,5,7)--(6,5,6);
    \draw[line width=2pt, gray](10,5,7)--(5,0,6);
    \draw[line width=2pt, gray](1,0,5)--(5,0,6);
    \draw[line width=2pt, gray](6,5,6)--(5,5,1);
    \draw[line width=2pt, gray](9,5,2)--(5,5,1);
    \draw[line width=2pt, gray](10,5,7)--(9,5,2);
     
    \draw[->, line width=2pt,red, -stealth](0,0,0)--(1,0,5)node[below left]{$\vec{u}$};
    \draw[->, line width=2pt, blue, -stealth](0,0,0)--(5,5,1)node[above left]{$\vec{w}$} ;
    \node[blue] at (0.4, 1.1)   (b) {$\theta$};
    \draw[->, line width=1pt, -stealth](0,0,0)--(0,6,0) node[above]{$\vec{u}\times\vec{v}$};
    \draw[->, gray, line width=0.5pt, -stealth](-0.25,2.5,0)node[left, black]{$h$}--(-0.25,0,0);
    \draw[->, gray, line width=0.5pt, -stealth](-0.25,2.5,0)--(-0.25,5,0);
\end{tikzpicture}
\end{center}
 
The height of the parallelepiped is measured along a line perpendicular to the base.  But, since the height is perpendicular to the base, it is thus normal to the base, so we know that $\vec{u}\times\vec{v}$ lies on such a line!

Let $\theta$ be the angle between $\vec{w}$ and $\vec{u}\times\vec{v}$, $0\leq \theta\leq\pi$.  Then the height, $h$, of the parallelepiped is given by
$$h=\norm{\vec{w}}|\cos\theta |$$
 
It may be difficult to visualize this in two dimensions.  Below is a replica of of the above diagram in GeoGebra.  RIGHT-CLICK and DRAG to rotate the image.
 
\begin{onlineOnly}
\begin{center}
\geogebra{tfuzeqwr}{900}{700}
\end{center}
\end{onlineOnly}
 
Hopefully the use of cosine looks familiar, indeed the dot product of two vectors $\vec{u}\cdot\vec{v}$ relates exactly in this way to $\cos(\theta)!$

This gives us the following formula for the volume of the parallelepiped
$$\mbox{Volume}=\norm{\vec{u}\times\vec{v}}\norm{\vec{w}}|\cos\theta |=|(\vec{u}\times\vec{v})\dotp\vec{w}|$$
 
We have established the following formula.
 
\begin{formula}\label{form:volumeparallelepiped}
The volume of a parallelepiped determined by vectors $\vec{u}$, $\vec{v}$ and $\vec{w}$ in $\RR^3$ is given by\\
$$\mbox{Volume}=|(\vec{u}\times\vec{v})\dotp\vec{w}|$$
\end{formula}
 
In fact, this aligns with our characterization of volume-measurement via the determinant of a $3\times 3$ matrix whose columns are the vectors that determine the parallelepiped.
 
\begin{formula}\label{form:boxproduct}
Let $\vec{u}=\begin{bmatrix}u_1\\u_2\\u_3\end{bmatrix},\quad\vec{v}=\begin{bmatrix}v_1\\v_2\\v_3\end{bmatrix},\quad\vec{w}=\begin{bmatrix}w_1\\w_2\\w_3\end{bmatrix}$ be vectors in $\RR^3$.  Then the volume of the parallelepiped determined by $\vec{u}$, $\vec{v}$ and $\vec{w}$ is given by
$$\mbox{Volume}=\Big|(\vec{u}\times\vec{v})\dotp\vec{w}\Big|=\Big|\det\begin{bmatrix}w_1&u_1&v_1\\w_2&u_2&v_2\\w_3&u_3&v_3\end{bmatrix}\Big|$$

This is sometimes referred to as the \dfn{box product} or the \dfn{scalar triple product}.
\end{formula}

\end{document}