\documentclass{ximera}
\input{../../../preamble.tex}

\author{Zack Reed}
%borrowed from selinger linear algebra
\title{Angles and the Dot Product}
\begin{document}
\begin{abstract}

\end{abstract}
\maketitle

\section*{Dot Product and Angle}
 
Given two vectors $\vec{u}$ and $\vec{v}$, let $\theta$ be the angle between them such that $0\leq\theta\leq \pi$.  We will refer to $\theta$ as the \dfn{included angle}.
 
\begin{center}
\begin{tikzpicture}
\node[] at (0.8, 0.2)   (a) {$\theta$};
\draw (0.5,0) arc (0:30:0.5) ;
\draw[line width=1pt,-stealth,red](0,0)--(2.5,0)node[below right]{$\vec{u}$};
\draw[line width=1pt,-stealth, blue](0,0)--(3.46,2)node[above right]{$\vec{v}$};
 \end{tikzpicture}
\end{center}
 
The following theorem establishes a relationship between the dot product and the included angle.
 
  \begin{theorem}\label{th:dotproductcosine} Let $\vec{u}$ and $\vec{v}$ be vectors in $\RR^n$, and let $\theta$ be the included angle.  Then
  \begin{equation*} \label{matlintrans}
 \vec{u}\dotp\vec{v}=\norm{\vec{u}}\norm{\vec{v}}\cos \theta
\end{equation*}
\end{theorem}
 
 
\begin{proof} Consider the triangle formed by $\vec{u}$, $\vec{v}$ and $\vec{u}-\vec{v}$.
 
\begin{center}
\begin{tikzpicture}
\node[] at (0.8, 0.2)   (a) {$\theta$};
\draw (0.5,0) arc (0:30:0.5) ;
\draw[line width=1pt,-stealth,red](0,0)--(2.5,0)node[below right]{$\vec{u}$};
\draw[line width=1pt,-stealth, blue](0,0)--(3.46,2)node[above right]{$\vec{v}$};
\draw[line width=1pt,-stealth](3.46,2)--(2.5,0);
\node[] at (3.5, 0.9)   (a) {$\vec{u}-\vec{v}$};
 \end{tikzpicture}
\end{center}
 
By the Law of Cosines we have:
\begin{align*}
\norm{\vec{u}-\vec{v}}^2=\norm{\vec{u}}^2+\norm{\vec{v}}^2-2\norm{\vec{u}}\norm{\vec{v}}\cos\theta
\end{align*}
By the properties of dot products from \ref{th:dotproductproperties}\ref{item:norm} of \href{https://ximera.osu.edu/appliedlinearalgebra/c5ChapterFive/learningActivities/m5LearningActivities/m5DotProduct/dotProduct}{before} we can do the following:
\begin{align*}
(\vec{u}-\vec{v})\dotp (\vec{u}-\vec{v})=&\vec{u}\dotp \vec{u}+\vec{v}\dotp \vec{v}-2\norm{\vec{u}}\norm{\vec{v}}\cos\theta
\end{align*}

\begin{align*}
(\vec{u}-\vec{v})\dotp \vec{u}-(\vec{u}-\vec{v})\dotp \vec{v}=&\vec{u}\dotp \vec{u}+\vec{v}\dotp \vec{v}-2\norm{\vec{u}}\norm{\vec{v}}\cos\theta
\end{align*}

\begin{align*}
\vec{u}\dotp \vec{u}-\vec{v}\dotp\vec{u}-\vec{u}\dotp\vec{v}+\vec{v}\dotp \vec{v}=&\vec{u}\dotp \vec{u}+\vec{v}\dotp \vec{v}-2\norm{\vec{u}}\norm{\vec{v}}\cos\theta
\end{align*}

\begin{align*}
-2(\vec{u}\dotp \vec{v})=&-2\norm{\vec{u}}\norm{\vec{v}}\cos\theta\\
\vec{u}\dotp \vec{v}=&\norm{\vec{u}}\norm{\vec{v}}\cos\theta
\end{align*}
\end{proof}
 
\begin{example}\label{ex:anglebetweenvectors}
Find the included angle between vectors $\vec{u}=\begin{bmatrix}2\\-1\\4\end{bmatrix}$ and $\vec{v}=\begin{bmatrix}1\\2\\-3\end{bmatrix}$.
 
\begin{explanation}
By Theorem \ref{th:dotproductcosine}, $\cos \theta=\frac{\vec{u}\dotp \vec{v}}{\norm{\vec{u}}\norm{\vec{v}}}$. 
$$\vec{u}\dotp\vec{v}=2-2-12=-12$$
$$\norm{\vec{u}}=\sqrt{4+1+16}=\sqrt{21}$$
$$\norm{\vec{v}}=\sqrt{1+4+9}=\sqrt{14}$$
$$\cos\theta =\frac{-12}{\sqrt{21}\sqrt{14}}$$
$$\theta \approx 134.4^{\circ}  $$
\end{explanation}
\end{example}
 
\subsection*{Orthogonal Vectors}
 
\begin{definition}\label{def:orthovectors}
Let $\vec{u}$ and $\vec{v}$ be vectors in $\RR^n$. We say $\vec{u}$ and $\vec{v}$ are \dfn{orthogonal} if $\vec{u}\dotp \vec{v}=0$.
\end{definition}
 
We can use Theorem \ref{th:dotproductcosine} to show that two non-zero \dfn{orthogonal} vectors of $\RR^n$ are simply \dfn{perpendicular} vectors (the included angle is $90^{\circ}$).  To see this, suppose that $\vec{u}\dotp\vec{v}=0$ for nonzero vectors $\vec{u},\vec{v}$.  Then from Theorem \ref{th:dotproductcosine} we have
$$0=\norm{\vec{u}}\norm{\vec{v}}\cos\theta.$$
Since $\vec{u},\vec{v}$ are nonzero vectors, we have $0=\cos\theta$, which implies $\theta=90^{\circ}$.  The converse also holds.  If $\theta=90^{\circ}$, then the dot product is clearly 0.
 
The reason we prefer the term ``orthogonal" to ``perpendicular" in this course is because $\RR^n$ is only one example of a \dfn{vector space}, and the dot product is only one example of a more general product, called an \dfn{inner product}.  For vectors in $\RR^n$ a zero dot product happens to coincide with the geometric idea of perpendicularity, but there are many vector spaces that do not possess the visual geometry of $\RR^n$.  (Later in the text, you will encounter vector spaces whose vectors are polynomial functions!)  In these more abstract settings, a zero inner product still signals a special relationship between vectors.  The term \dfn{orthogonal} captures this relationship.
 
\begin{remark}  By our definition, the zero vector is orthogonal to any vector.  However, we will not use the word \dfn{perpendicular} when the zero vector is involved, as it is not possible to talk about an ``included angle''.
\end{remark}
 
%For $\RR^n$ with $n&gt;3$, when we no longer have visual interpretation to aid us, our definition of \dfn{orthogonal} vectors in terms of dot product will come in handy.
 
\section*{Practice Problems}
\begin{problem}%\label{prob:anglebetweenvectors}
Find the degree measure of the included angle, $\theta$ for each pair of vectors.  Round your answers to the nearest tenth.
  \begin{problem}\label{prob:anglebetweenvectors1}
  $\begin{bmatrix}1\\2\end{bmatrix}$ and $\begin{bmatrix}-3\\-1\end{bmatrix}$.
   
  Answer: $\theta=\answer{135}^\circ$
  \end{problem}
   
  \begin{problem}\label{prob:anglebetweenvectors2}
  $\begin{bmatrix}-1\\2\\4\end{bmatrix}$ and $\begin{bmatrix}-2\\1\\-1\end{bmatrix}$
   
   Answer: $\theta=\answer{90}^\circ$
  \end{problem}
 
\begin{problem}\label{prob:anglebetweenvectors3}
  $\begin{bmatrix}0\\-3\\1\end{bmatrix}$ and $\begin{bmatrix}-5\\-2\\4\end{bmatrix}$
   
   Answer: $\theta=\answer[tolerance=0.1]{61.9}^\circ$
  \end{problem}
 
\begin{problem}\label{prob:anglebetweenvectors4}
  $\begin{bmatrix}1\\1\\-1\\2\end{bmatrix}$ and $\begin{bmatrix}-2\\0\\-3\\1\end{bmatrix}$
   
   Answer: $\theta=\answer[tolerance=0.1]{72.4}^\circ$
  \end{problem}
   
  \end{problem}
 
\begin{problem}\label{prob:dotproductsign}
What does the sign of the dot product tell us about the relative orientation of the vectors?
\begin{multipleChoice}
  \choice{Whether one vector is above another.}
  \choice{Whether one vector is smaller or larger than another.}
  \choice[correct]{Whether the angles are oriented together (pointing in the same direction) or away from each other (pointing in the opposite direction)}
\end{multipleChoice}
\end{problem}
 
\begin{problem}\label{prob:orthvectorsdot}
Find all values of $a$ so that $\begin{bmatrix}a^2\\2a\\1\end{bmatrix}$ is orthogonal to $\begin{bmatrix}1\\2\\3\end{bmatrix}$.  List your answers in increasing order.
 
Answer: $\answer{-3}, \answer{-1}$.
\end{problem}
 
\begin{problem}\label{prob:parallelvectdot}
Find the value of $x$ for which the vector $\begin{bmatrix}x\\-4\end{bmatrix}$ is parallel to the vector $\begin{bmatrix}3\\2\end{bmatrix}$. What is the measure of the included angle, $\theta$? Find the measure of the included angle using Theorem \ref{ex:anglebetweenvectors}.  Do the two results agree?
 
Answer:
$$x=\answer{-6}$$
$$\theta=\answer{180}^\circ$$
\end{problem}


\end{document}