\documentclass{ximera}
\input{../../preamble.tex}

\author{Zack Reed}
%borrowed from selinger linear algebra
\begin{document}


\begin{exercise}

    Which of the following vectors are redundant? If there are redundant
vectors, write each of them as a linear combination of previous
vectors.

(NOTE: Take the vectors in order. You want to consider the first available linearly independent vectors when filling in the answers.)

\begin{hint}

    If it's not immediately obvious which vectors are linear combinations of others,that's totally okay! 

    You'll want to use MATLAB to solve the linear combination equation for the redundant vectors.

    First, use \texttt{assume($a^2+b^2+c^2+d^2>0$)} to get rid of the trivial solution, then use \texttt{solve} to see if a linear combination of the vectors add to the zero vector.

    If you use the \texttt{"ReturnConditions",true} arguments at the end of solve, it will tell you which coefficients are non-zero. 

    You might need to do this more than once depending on which and how many vectors are redundant.

\end{hint}
Consider the following set of vectors in 4-dimensional space:
\[
\vect{v}_1 = \startmat{r} 1 \\ 0 \\ 2 \\ 1 \stopmat,\quad
\vect{v}_2 = \startmat{r} 0 \\ 1 \\ 1 \\ 3 \stopmat,\quad
\vect{v}_3 = \startmat{r} 1 \\ 1 \\ 3 \\ 4 \stopmat,\quad
\vect{v}_4 = \startmat{r} 2 \\ 2 \\ 7 \\ 10 \stopmat,\quad
\vect{v}_5 = \startmat{r} 5 \\ 1 \\ 13 \\ 12 \stopmat.
\]


Vectors \begin{selectAll}
\choice{$\vec{v}_1$}
\choice{$\vec{v}_2$}
\choice[correct]{$\vec{v}_3$}
\choice{$\vec{v}_4$}
\choice[correct]{$\vec{v}_5$}
\end{selectAll} are redundant.

We have
The first redundant vector is the made from the linear combination $\answer{1}\vect{v}_{\answer{1}}+\answer{1}\vect{v}_{\answer{2}}$ and the second redundant vector is made from the linear combination $\answer{1}\vect{v}_{\answer{1}}+\answer{-3}\vect{v}_{\answer{2}}+\answer{2}\vect{v}_{\answer{4}}$.


\end{exercise}

\end{document}