\documentclass{ximera}
\input{../../preamble.tex}

\author{Zack Reed}
%borrowed from selinger linear algebra
\begin{document}


\begin{exercise}

    Which of the following vectors are redundant? If there are redundant
vectors, write each of them as a linear combination of previous
vectors.

(NOTE: Take the vectors in order. You want to consider the first available linearly independent vectors when filling in the answers.)

\begin{hint}

    If it's not immediately obvious which vectors are linear combinations of others,that's totally okay! 
    
    You'll want to use MATLAB to solve the linear combination equation for the redundant vectors.

    First, use \texttt{assume(a\^2+b\^2+c\^2>0)} to get rid of the trivial solution, then use \texttt{solve} to see if a linear combination of the vectors add to the zero vector.

    If you use the \texttt{"ReturnConditions",true} arguments at the end of solve, it will tell you which coefficients are non-zero. 

    You might need to do this more than once depending on which and how many vectors are redundant.

\end{hint}
\begin{equation*}
  \vect{u}_1 = \startmat{r} 1 \\ 0 \\ 1 \stopmat,\quad
  \vect{u}_2 = \startmat{r} 2 \\ 0 \\ 2 \stopmat,\quad
  \vect{u}_3 = \startmat{r} 1 \\ 2 \\ 1 \stopmat,\quad
  \vect{u}_4 = \startmat{r} 1 \\ 6 \\ 1 \stopmat.
\end{equation*}

  Vectors $\answer{2}$ and $\answer{4}$ are redundant. (Say the number, so vector $\vec{u}_3$ is vector $3$.)
  
  We have


  $\text{Vector }\answer{2}=\answer{2}\text{ times vector }\answer{1}$ and 
  
  $\text{Vector }\answer{4}=\answer{3}\text{ times vector }\answer{3}-\answer{2}\text{ times vector }\answer{1}$.


\end{exercise}

\end{document}