\documentclass{ximera}
\input{../../preamble.tex}

\author{Zack Reed}
%borrowed from selinger linear algebra
\begin{document}


\begin{exercise}

    True or false? Explain.
    \begin{enumerate}
    \item Every set of $5$ vectors in $\R^5$ \wordChoice{
        \choice{must be}
        \choice[correct]{could be}
        \choice{cannot be}
    } linearly independent.
        \begin{problem}
            The 
        \end{problem}
    \item Every set of $4$ vectors in $\R^5$ is linearly independent.
    \item Every set of $6$ vectors in $\R^5$ is linearly dependent.
    \item No set of $4$ vectors spans $\R^5$.
    \item Every linearly independent set of $5$ vectors in $\R^5$ is a
      basis of $\R^5$.
    \item Every linearly independent set of $4$ vectors in $\R^5$ is a
      basis of $\R^5$.
    \item Some linearly independent set of $4$ vectors in $\R^5$ is a
      basis of $\R^5$.
    \item Every spanning set of $6$ vectors in $\R^5$ is a basis of
      $\R^5$.
    \item Every linearly independent set of $4$ vectors in $\R^5$ spans
      a $4$-dimensional subspace of $\R^5$.
    \end{enumerate}

      \begin{enumerate}
      \item No. For example, the vectors
        $\set{\vect{0},\vect{0},\vect{0},\vect{0},\vect{0}}$ are
        linearly dependent.
      \item No. For example, the vectors
        $\set{\vect{0},\vect{0},\vect{0},\vect{0}}$ are linearly
        dependent.
      \item Yes, by Proposition~\ref{prop:size-linearly-independent-or-spanning}.
      \item Correct, by Proposition~\ref{prop:size-linearly-independent-or-spanning}.
      \item Yes, by Proposition~\ref{prop:basis-test-k-vectors}.
      \item No, in fact no such set is a basis of $\R^5$, since every
        basis of $\R^5$ consists of $5$ vectors by
        Theorem~\ref{thm:bases-same-size}.
      \item No, as noted in the previous answer.
      \item No, no basis of $\R^5$ can have $6$ elements.
      \item Yes, because a linearly independent set of vectors is a
        basis of the subspace it spans.
      \end{enumerate}

\end{exercise}

\end{document}