\documentclass{ximera}
\input{../../preamble.tex}

\author{Zack Reed}
%borrowed from selinger linear algebra
\begin{document}


\begin{exercise}

    Consider the following augmented matrices, where $h$ and $k$ are arbitrary numbers. Find values for $h$ and $k$ that satisfy the given conditions.

    If no such values exist, enter "none." If any value of $h$ or $k$ will work, enter "any."

    \begin{enumerate}
        \item 
        \begin{equation*}
          \startmat{rr|r}
            2 & h & 4 \\
            3 & 6 & 7
          \stopmat
        \end{equation*}
  
        The system is inconsistent when $h = \answer{4}$.

        If $h\neq 4$, the system is consistent when $h = \answer[format=string]{any}$.

        \begin{hint}
        
            Find the rref of the matrix and determine for which values of h the matrix will have definite values. Any value for which the rref would not be defined would make the system inconsistent.

            Test this out for any inconsistent values! Plug it in and see what the resulting rref matrix looks like.


        \end{hint}
        
        
        \item 
        \begin{equation*}
          \startmat{rr|r}
            1 & h & 3 \\
            2 & 4 & 6
          \stopmat
        \end{equation*}

        The system is inconsistent when $h = \answer[format=string]{none}$. The system is consistent when $h = \answer[format=string]{any}$.
        
        
        \item Choose $h$ and $k$ such that the augmented matrix shown has each of the following:
        \begin{enumerate}
          \item one solution
          \item no solution
          \item infinitely many solutions
        \end{enumerate}
        \begin{equation*}
          \startmat{rr|r}
            1 & h & 2 \\
            2 & 4 & k
          \stopmat
        \end{equation*}
        
        If $h\neq \answer{2}$ there will be a unique solution for $k=\answer[format=string]{any}$. If $h=\answer{2}$ and $k\neq \answer{4}$, there are no solutions. If $h=\answer{2}$ and $k=\answer{4}$, then there are infinitely many solutions.

        \begin{hint}
        
            Set $h$ and $k$ to symbolic variables and find the rref of the matrix. Once you determine the relevant information about $h$, set $h$ to the inconsistent value and test out different values of $k$ to see when (or if) the system becomes consistent.

        \end{hint}
        
        \item Choose $h$ and $k$ such that the augmented matrix shown has each of the following:
        \begin{enumerate}
          \item one solution
          \item no solution
          \item infinitely many solutions
        \end{enumerate}
        \begin{equation*}
          \startmat{rr|r}
            1 & 2 & 2 \\
            2 & h & k
          \stopmat
        \end{equation*}
        
        If $h\neq \answer{4}$, then there is exactly one solution. If $h=\answer{4}$ and $k\neq \answer{4}$, then there are no solutions. If $h=\answer{4}$ and $k=\answer{4}$, then there are \wordChoice{
            \choice{no}
            \choice{exactly one}
            \choice[correct]{infinitely many}} solutions.
        
    \end{enumerate}
    

\end{exercise}

\end{document}