\documentclass{ximera}
\input{../../preamble.tex}

\author{Zack Reed}
%borrowed from selinger linear algebra
\begin{document}

\begin{exercise}
  For each of the following matrices, find the eigenvectors and
  eigenvalues. Use this to visualize the linear transformation
  $T:\R^2\to\R^2$ that is described by the matrix.
  \begin{equation*}
    A = \startmat{rr}
      1 & 2 \\
      2 & 1 \\
    \stopmat,
    \quad
    B = \startmat{rr}
      1 & 0 \\
      2 & -1 \\
    \stopmat,
    \quad
    C = \startmat{rr}
      1 & 1 \\
      0 & 2 \\
    \stopmat,
    \quad
    D = \startmat{rr}
      1 & 1 \\
      1 & 1 \\
    \stopmat,
    \quad
    E = \startmat{rr}
      1 & 0 \\
      -1 & 2 \\
    \stopmat.
  \end{equation*}
\end{exercise}

\begin{exercise}
  If $A$ is the matrix of a linear transformation that rotates all
  vectors in $\R^2$ by $60^{\circ}$, explain why $A$ cannot have any
  real eigenvalues. Is there an angle such that rotation by this
  angle would have a real eigenvalue? What eigenvalues would be
  obtainable in this way?
  \begin{solution}
    A rotation by $60^{\circ}$ cannot map any non-zero vector to a
    scalar multiple of itself. The only rotations of $\R^2$ that have
    real eigenvalues are rotations by $180^{\circ}$ and $0^{\circ}$.
    The former has eigenvalue $-1$, and the latter has eigenvalue $1$
    (with all vectors being eigenvectors).
  \end{solution}
\end{exercise}

\begin{exercise}
  Let $T$ be the linear transformation that reflects vectors about
  the $x$-axis. Find a matrix for $T$ and then find its eigenvalues and
  eigenvectors.
  \begin{solution}
    The matrix of $T$ is $\startmat{rr}
      1 & 0 \\
      0 & -1
    \stopmat$. The eigenvectors and eigenvalues are:
    \begin{equation*}
      \set{\startmat{c}
          0 \\
          1
        \stopmat} ~\mbox{for eigenvalue}~ {-1},
      \quad\set{\startmat{c}
          1 \\
          0
        \stopmat} ~\mbox{for eigenvalue}~ {1}.
    \end{equation*}
  \end{solution}
\end{exercise}

\begin{exercise}
  Let $T$ be the linear transformation that reflects vectors about the
  line $x=y$. Find a matrix of $T$ and then find eigenvalues and
  eigenvectors.
  \begin{solution}
    The matrix of $T$ is $\startmat{rr}
      0 & 1 \\
      1 & 0
    \stopmat$. The eigenvectors and eigenvalues are:
    \begin{equation*}
      \set{\startmat{c}
          1 \\
          1
        \stopmat} ~\mbox{for eigenvalue}~ {-1},
      \quad\set{\startmat{c}
          -1 \\
          1
        \stopmat} ~\mbox{for eigenvalue}~ {1}.
    \end{equation*}
  \end{solution}
\end{exercise}

\begin{exercise}
  Let $T$ be the linear transformation that reflects all vectors in
  $\R^3$ about the $xy$-plane. Find a matrix for $T$ and then
  obtain its eigenvalues and eigenvectors.
  \begin{solution}
    The matrix of $T$ is $\startmat{rrr}
      1 & 0 & 0 \\
      0 & 1 & 0 \\
      0 & 0 & -1
    \stopmat$
    The eigenvectors and eigenvalues are:
    \begin{equation*}
      \set{\startmat{c}
          0 \\
          0 \\
          1
        \stopmat} ~\mbox{for eigenvalue $-1$},
      \quad\set{\startmat{c}
          1 \\
          0 \\
          0
        \stopmat,
        \startmat{c}
          0 \\
          1 \\
          0
        \stopmat} ~\mbox{for eigenvalue $1$}.
    \end{equation*}
  \end{solution}
\end{exercise}

\end{document}